---
prefer-html: true
---

# Common Statistical Tests

```{r}
#| label: setup-statistical-tests
#| include: false
# Load required packages for data analysis and visualization
library(tidyverse)    # Core tidyverse (includes ggplot2, dplyr, tidyr, readr, etc.)
library(rstatix)      # Pipe-friendly statistical tests
library(knitr)        # Document generation
library(kableExtra)   # Enhanced tables
library(broom)        # Tidy model outputs
library(car)          # Companion to Applied Regression (for Levene's test)

# Set global knitr options
knitr::opts_chunk$set(

  echo = TRUE,

  warning = FALSE,

  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  comment = "#>"
)

# Set a consistent theme for all plots
theme_set(theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40"),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom"
  ))
```

## Introduction

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

1.  Select appropriate statistical tests based on data type, distribution, and research question
2.  Perform and interpret parametric tests (t-tests, ANOVA, Pearson correlation)
3.  Apply non-parametric alternatives (Mann-Whitney U, Kruskal-Wallis, Spearman correlation) when assumptions are violated
4.  Conduct post-hoc analyses to identify specific group differences
5.  Fit and interpret linear and multiple regression models
6.  Report statistical results in a clear, standard format
:::

Chapter 4 set up the logic of testing. This chapter goes through the actual tests you will reach for in practice. T-tests when you have two groups. ANOVA when you have more. Non-parametric versions of both when the data refuse to behave. Correlation and regression when you care about relationships rather than differences. By the end you should be able to pick the right test for a given question and run it without second-guessing yourself.

## Choosing the right statistical test

There is no single answer to "which test should I use?". It depends on a few things, and once you can answer them, the choice is usually obvious:

1. What are you actually asking? Difference between groups, relationship between variables, change over time?
2. What kind of data do you have? Continuous, categorical, ordinal?
3. How many groups or variables are involved?
4. Are the data roughly normal, or skewed?
5. Are your observations independent, or paired/repeated?

### Decision Tree for Common Tests

```{r}
#| label: fig-decision-tree
#| echo: false
#| fig-cap: "Decision tree for selecting appropriate statistical tests"
#| fig-alt: "DiagrammeR/Graphviz decision tree starting from 'Research Question' and branching by number of variables and sample structure to recommend specific tests (t-tests, Wilcoxon, ANOVA, Kruskal-Wallis, correlation, regression, MANOVA, PCA, clustering)."
#| eval: false

# This code requires the DiagrammeR package
# If you want to run this code, first install the package with:
# install.packages("DiagrammeR")

library(DiagrammeR)

grViz("
digraph test_selection {
  # Node definitions
  node [shape = rectangle, fontname = 'Arial', fontsize = 10, style = filled, fillcolor = lightblue]

  start [label = 'Research Question', fillcolor = lightgreen]

  # First level: Number of variables
  one_var [label = 'One Variable']
  two_var [label = 'Two Variables']
  multi_var [label = 'Multiple Variables']

  # Second level for one variable
  one_sample [label = 'One Sample']
  two_sample [label = 'Two Samples']
  multi_sample [label = 'Multiple Samples']

  # Tests for one variable
  t_test [label = 'One-Sample t-Test\\n(Normal, Continuous)']
  wilcox [label = 'Wilcoxon Signed-Rank\\n(Non-normal, Continuous)']
  binom [label = 'Binomial Test\\n(Categorical)']

  # Tests for two variables - two samples
  ind_t [label = 'Independent t-Test\\n(Normal, Continuous)']
  paired_t [label = 'Paired t-Test\\n(Normal, Continuous, Related)']
  mann [label = 'Mann-Whitney U\\n(Non-normal, Continuous)']
  wilcox_paired [label = 'Wilcoxon Signed-Rank\\n(Non-normal, Continuous, Related)']
  chi_sq [label = 'Chi-Square Test\\n(Categorical)']
  mcnemar [label = 'McNemar Test\\n(Categorical, Related)']

  # Tests for multiple samples
  anova [label = 'ANOVA\\n(Normal, Continuous)']
  rm_anova [label = 'Repeated Measures ANOVA\\n(Normal, Continuous, Related)']
  kruskal [label = 'Kruskal-Wallis\\n(Non-normal, Continuous)']
  friedman [label = 'Friedman Test\\n(Non-normal, Continuous, Related)']
  chi_sq_multi [label = 'Chi-Square Test\\n(Categorical)']

  # Tests for relationships
  pearson [label = 'Pearson Correlation\\n(Normal, Continuous)']
  spearman [label = 'Spearman Correlation\\n(Non-normal or Ordinal)']
  linear_reg [label = 'Linear Regression\\n(Continuous Predictor & Outcome)']
  logistic [label = 'Logistic Regression\\n(Continuous Predictor, Binary Outcome)']

  # Multiple variables
  manova [label = 'MANOVA\\n(Multiple Continuous Outcomes)']
  pca [label = 'Principal Component Analysis\\n(Dimension Reduction)']
  cluster [label = 'Cluster Analysis\\n(Grouping)']

  # Connections
  start -> {one_var two_var multi_var}

  one_var -> {one_sample two_sample multi_sample}

  one_sample -> {t_test wilcox binom}

  two_sample -> {ind_t paired_t mann wilcox_paired chi_sq mcnemar}

  multi_sample -> {anova rm_anova kruskal friedman chi_sq_multi}

  two_var -> {pearson spearman linear_reg logistic}

  multi_var -> {manova pca cluster}
}
")
```

::: {.callout-note}
## Code Explanation

This code creates a decision tree for statistical test selection:

1. **Package Setup**:
   - Uses `DiagrammeR` for creating flowcharts
   - Defines node styles and attributes
   - Sets up the graph structure

2. **Node Structure**:
   - Main categories: One Variable, Two Variables, Multiple Variables
   - Subcategories for different data types
   - Specific tests for each scenario

3. **Connections**:
   - Shows logical flow between decisions
   - Links tests to appropriate scenarios
   - Guides test selection process
:::

::: {.callout-important}
## Results Interpretation

The decision tree in @fig-decision-tree provides a systematic approach to test selection:

1. **Test Categories**:
   - Parametric vs. non-parametric tests
   - Tests for different data types
   - Tests for different sample sizes

2. **Selection Criteria**:
   - Data distribution (normal vs. non-normal)
   - Number of variables
   - Type of variables (continuous vs. categorical)
   - Sample independence

3. **Test Properties**:
   - Assumptions of each test
   - Appropriate use cases
   - Limitations and considerations
:::

::: {.callout-tip}
## PROFESSIONAL TIP: Statistical Test Selection

When choosing statistical tests:

1. **Data Characteristics**:
   - Check data distribution
   - Verify assumptions
   - Consider sample size
   - Evaluate data types

2. **Research Design**:
   - Match test to research question
   - Consider experimental design
   - Account for dependencies
   - Plan for multiple comparisons

3. **Test Selection Process**:
   - Start with research question
   - Identify data characteristics
   - Choose appropriate test
   - Verify assumptions
   - Consider alternatives
:::

## Parametric vs. Non-Parametric Tests

### Parametric Tests

Parametric tests make assumptions about the underlying population distribution, typically that the data follows a normal distribution. Common parametric tests include:

- t-tests
- ANOVA
- Pearson correlation
- Linear regression

### Non-Parametric Tests

Non-parametric tests make fewer assumptions about the population distribution and are useful when data doesn't meet the assumptions of parametric tests. Common non-parametric tests include:

- Mann-Whitney U test
- Wilcoxon signed-rank test
- Kruskal-Wallis test
- Spearman correlation

### Checking Assumptions

Before applying a parametric test, it's essential to check if your data meets the necessary assumptions. Let's use our crop yield dataset to demonstrate:

```{r}
#| label: load-crop-data
#| echo: true

# Packages are loaded in the chapter setup chunk above.

# Load the crop yield dataset
crop_yields <- read_csv("../data/agriculture/crop_yields.csv")

# Extract wheat yields for analysis
wheat_yields <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & Year >= 1960) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`)
```

```{r}
#| label: tbl-wheat-sample
#| tbl-cap: "Sample of Wheat Yield Data"

# View the first few rows
knitr::kable(head(wheat_yields),
             align = c("l", "c", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-normality-check
#| fig-cap: "Normality assessment for wheat yields"
#| fig-alt: "Two-panel base-R figure: left panel is a histogram of wheat yields (tonnes/ha) with a right-skewed shape; right panel is a normal Q-Q plot with a red reference line showing departures from normality in both tails."

# Check for normality - Visual methods
par(mfrow = c(1, 2))
hist(wheat_yields$`Wheat (tonnes per hectare)`, main = "Histogram of Wheat Yields", xlab = "Yield (tonnes/hectare)")
qqnorm(wheat_yields$`Wheat (tonnes per hectare)`); qqline(wheat_yields$`Wheat (tonnes per hectare)`, col = "red")
```

```{r}
#| label: tbl-shapiro-test
#| tbl-cap: "Shapiro-Wilk Normality Test Results: Wheat Yields"

set.seed(42)
# Statistical test for normality
shapiro_result <- shapiro.test(sample(wheat_yields$`Wheat (tonnes per hectare)`, min(5000, length(wheat_yields$`Wheat (tonnes per hectare)`))))

# Create a formatted table of the results
shapiro_table <- data.frame(
  Statistic = c("W-value", "p-value"),
  Value = c(
    round(shapiro_result$statistic, 2),
    format.pval(shapiro_result$p.value, digits = 3)
  )
)

# Display the formatted table
knitr::kable(shapiro_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-wheat-summary
#| tbl-cap: "Summary Statistics: Wheat Yields"

# Summary statistics
summary_stats <- wheat_yields %>%
  summarize(
    n = n(),
    Mean = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    SD = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Min = round(min(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Max = round(max(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(summary_stats,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to check the normality assumption for parametric tests:

1. **Data Preparation**: Imports and filters the dataset, removing missing values
2. **Visual Assessment**: Creates histogram and Q-Q plot to visually assess normality
3. **Statistical Testing**: Uses Shapiro-Wilk test to formally test for normality
4. **Result Presentation**: Formats results in clear, publication-ready tables
:::

::: {.callout-important}
## Results Interpretation
The normality assessment in @fig-normality-check reveals a right-skewed distribution of wheat yields, confirmed by both visual inspection and the significant Shapiro-Wilk test (@tbl-shapiro-test, p < 0.001). This suggests non-parametric tests may be more appropriate for these data, or that transformations should be considered before applying parametric methods.
:::

## Tests for Comparing Groups

### t-Tests

#### Independent Samples t-Test

Used to compare means between two independent groups. Let's compare wheat yields between two time periods:

```{r}
#| label: independent-ttest-data
#| echo: true

# Create two groups: early period (before 2000) and recent period (2000 onwards)
crop_yields_grouped <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & Year >= 1960) %>%
  mutate(period = ifelse(Year < 2000, "Early Period (pre-2000)", "Recent Period (2000+)"))
```

```{r}
#| label: fig-independent-ttest
#| fig-cap: "Wheat yields by time period"
#| fig-alt: "Two-category boxplot of wheat yields (tonnes/ha) comparing the early period (pre-2000) to the recent period (2000+); the recent-period box sits noticeably higher, suggesting yields have increased over time."

# Visualize the data
ggplot(crop_yields_grouped, aes(x = period, y = `Wheat (tonnes per hectare)`, fill = period)) +
  geom_boxplot() +
  labs(title = "Wheat Yields by Time Period",
       x = "Period",
       y = "Wheat Yield (tonnes/hectare)") +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.callout-note}
## Checking t-test assumptions

Before the test, verify normality of each group and equality of variance.
`t.test()` defaults to Welch's correction, which is robust to unequal
variance — preferred over the classical Student's t-test in almost all
applied work.

```{r}
#| label: ttest-assumptions
by_period <- split(crop_yields_grouped$`Wheat (tonnes per hectare)`,
                   crop_yields_grouped$period)
shapiro.test(by_period[[1]])  # group 1 normality
shapiro.test(by_period[[2]])  # group 2 normality
var.test(`Wheat (tonnes per hectare)` ~ period, data = crop_yields_grouped)
```

If Shapiro rejects on either group at moderate sample size, prefer
`wilcox.test()` (Mann-Whitney U).
:::

```{r}
#| label: tbl-independent-ttest
#| tbl-cap: "Independent Samples t-Test Results: Wheat Yields by Time Period"

# Perform independent samples t-test using formula interface with backticks
t_test_result <- t.test(`Wheat (tonnes per hectare)` ~ period, data = crop_yields_grouped)

# Create a formatted table of the results
t_test_table <- data.frame(
  Statistic = c("t-value", "Degrees of Freedom", "p-value", "Mean Difference", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(t_test_result$statistic, 3),
    round(t_test_result$parameter, 1),
    format.pval(t_test_result$p.value, digits = 3),
    round(diff(t_test_result$estimate), 2),
    round(t_test_result$conf.int[1], 2),
    round(t_test_result$conf.int[2], 2)
  )
)

# Display the formatted table
knitr::kable(t_test_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-period-summary
#| tbl-cap: "Summary Statistics: Wheat Yields by Time Period"

# Summary statistics by period
period_summary <- crop_yields_grouped %>%
  group_by(period) %>%
  summarize(
    n = n(),
    Mean = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    SD = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Min = round(min(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Max = round(max(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(period_summary,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform an independent samples t-test:

1. **Data Preparation**: Creates two groups based on time periods
2. **Visualization**: Uses boxplots to visually compare the distributions
3. **Statistical Testing**: Performs t-test using R's formula interface with backticks for column names with spaces
4. **Result Presentation**: Creates formatted tables of results and summary statistics
:::

::: {.callout-important}
## Results Interpretation
The t-test results show a significant difference in wheat yields between the two time periods (p < 0.001). The recent period (2000+) has substantially higher yields (mean = 3.31 tonnes/hectare) compared to the earlier period (mean = 2.45 tonnes/hectare).

This increase of approximately 35% over time likely reflects advancements in agricultural technology, improved crop varieties, and better farming practices that have been developed and implemented over recent decades.
:::

#### Paired Samples t-Test

Used to compare means between two related groups. Let's compare wheat and rice yields for the same countries and years:

```{r}
#| label: paired-samples-data
#| echo: true

# Prepare data for paired t-test
paired_data <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & !is.na(`Rice (tonnes per hectare)`)) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`)
```

```{r}
#| label: tbl-paired-sample-data
#| tbl-cap: "Sample of Paired Crop Yield Data"

# View the first few rows
knitr::kable(head(paired_data),
             align = c("l", "c", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-paired-samples
#| fig-cap: "Comparison of wheat and rice yields"
#| fig-alt: "Two-category boxplot comparing wheat vs rice yields (tonnes/ha) from the paired country-year data; rice typically shows a higher and more spread-out distribution than wheat."

# Visualize the paired data
paired_data_long <- paired_data %>%
  pivot_longer(cols = c(`Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`), names_to = "Crop", values_to = "Yield")

ggplot(paired_data_long, aes(x = Crop, y = Yield, fill = Crop)) +
  geom_boxplot() +
  labs(title = "Comparison of Wheat and Rice Yields",
       x = "Crop Type",
       y = "Yield (tonnes/hectare)") +
  theme_minimal()
```

::: {.callout-note}
## Paired t-test assumption

The paired t-test assumes the *differences* (not the raw values) are
approximately normal. Check before running:

```{r}
#| label: paired-ttest-assumption
diffs <- paired_data$`Wheat (tonnes per hectare)` -
         paired_data$`Rice (tonnes per hectare)`
shapiro.test(diffs)
```

If the differences are heavily skewed, use `wilcox.test(..., paired = TRUE)`.
:::

```{r}
#| label: tbl-paired-ttest-results
#| tbl-cap: "Paired Samples t-Test Results: Wheat vs. Rice Yields"

# Perform paired t-test using vectors directly
paired_t_test <- t.test(
  paired_data$`Wheat (tonnes per hectare)`,
  paired_data$`Rice (tonnes per hectare)`,
  paired = TRUE
)

# Create a formatted table of the results
paired_t_test_table <- data.frame(
  Statistic = c("t-value", "Degrees of Freedom", "p-value", "Mean Difference", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(paired_t_test$statistic, 3),
    round(paired_t_test$parameter, 1),
    format.pval(paired_t_test$p.value, digits = 3),
    round(mean(paired_data$`Wheat (tonnes per hectare)` - paired_data$`Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    round(paired_t_test$conf.int[1], 2),
    round(paired_t_test$conf.int[2], 2)
  )
)

# Display the formatted table
knitr::kable(paired_t_test_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-paired-summary
#| tbl-cap: "Summary Statistics: Wheat vs. Rice Yields"

# Create a summary statistics table
paired_summary <- paired_data %>%
  summarize(
    n = n(),
    `Mean Wheat` = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    `SD Wheat` = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    `Mean Rice` = round(mean(`Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    `SD Rice` = round(sd(`Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    `Mean Difference` = round(mean(`Wheat (tonnes per hectare)` - `Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    `SD Difference` = round(sd(`Wheat (tonnes per hectare)` - `Rice (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(paired_summary,
             align = rep("r", 7),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform a paired samples t-test:

1. **Data Preparation**: Creates a dataset with paired observations (wheat and rice yields)
2. **Visualization**: Uses a scatterplot to show the relationship between the paired variables
3. **Statistical Testing**: Performs a paired t-test to compare the means of two related variables
4. **Result Presentation**: Creates formatted tables of results and summary statistics
:::

::: {.callout-important}
## Results Interpretation
The paired t-test results show a significant difference between wheat and rice yields (p < 0.001). On average, rice yields are higher than wheat yields by approximately 1.68 tonnes per hectare.

This difference is consistent with biological and agricultural knowledge - rice typically produces more biomass per hectare than wheat under certain conditions. The paired approach was appropriate here as it accounts for country-specific factors (climate, agricultural practices, economic development) that affect both crops similarly.
:::

### Analysis of Variance (ANOVA)

ANOVA is used to compare means among three or more independent groups. Let's compare crop yields across different continents:

```{r}
#| label: anova-data-setup
#| echo: true

# Create a mapping of countries to continents (simplified for demonstration)
continent_mapping <- tibble(
  Entity = c("United States", "Canada", "Mexico",
             "China", "India", "Japan",
             "Germany", "France", "United Kingdom",
             "Brazil", "Argentina", "Chile",
             "Egypt", "Nigeria", "South Africa",
             "Australia", "New Zealand"),
  Continent = c(rep("North America", 3),
                rep("Asia", 3),
                rep("Europe", 3),
                rep("South America", 3),
                rep("Africa", 3),
                rep("Oceania", 2))
)

# Join with crop yields data
continental_yields <- crop_yields %>%
  inner_join(continent_mapping, by = "Entity") %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & Year >= 2000)
```

```{r}
#| label: fig-anova-continents
#| fig-cap: "Wheat yields by continent (2000-present)"
#| fig-alt: "Boxplot of post-2000 wheat yields (tonnes/ha) grouped by continent and colored by continent; Europe shows the highest medians while Africa and Oceania sit lower, motivating an ANOVA comparison."

# Visualize wheat yields by continent
ggplot(continental_yields, aes(x = Continent, y = `Wheat (tonnes per hectare)`, fill = Continent)) +
  geom_boxplot() +
  labs(title = "Wheat Yields by Continent (2000-present)",
       x = "Continent",
       y = "Wheat Yield (tonnes/hectare)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Checking ANOVA assumptions

Before interpreting an ANOVA, verify that its assumptions hold:

1. **Independence** of observations (study design — not testable from the data alone).
2. **Normality** of residuals within each group (Shapiro-Wilk test or Q-Q plot).
3. **Homogeneity of variance** across groups (Levene's test).

If the assumptions are seriously violated, prefer the non-parametric **Kruskal-Wallis** test instead.

```{r}
#| label: anova-assumptions
#| message: false
#| warning: false

# Homogeneity of variance: Levene's test (car package)
if (requireNamespace("car", quietly = TRUE)) {
  car::leveneTest(`Wheat (tonnes per hectare)` ~ Continent, data = continental_yields)
}

# Per-group normality (Shapiro-Wilk on the largest groups)
continental_yields %>%
  group_by(Continent) %>%
  filter(n() >= 3 & n() <= 5000) %>%
  summarise(
    n = n(),
    shapiro_p = tryCatch(
      shapiro.test(`Wheat (tonnes per hectare)`)$p.value,
      error = function(e) NA_real_
    )
  )
```

A non-significant Levene's test (p > 0.05) supports the equal-variance assumption.
A non-significant Shapiro-Wilk (p > 0.05) supports normality within groups. With
large samples, Shapiro-Wilk becomes overly sensitive — also inspect a Q-Q plot
of the residuals: `plot(anova_result, which = 2)`.

```{r}
#| label: tbl-anova-results
#| tbl-cap: "ANOVA Results: Wheat Yields by Continent"

# Perform ANOVA
anova_result <- aov(`Wheat (tonnes per hectare)` ~ Continent, data = continental_yields)
anova_summary <- summary(anova_result)

# Create a formatted ANOVA table
anova_df <- data.frame(
  Source = c("Continent", "Residuals"),
  DF = c(anova_summary[[1]][["Df"]][1], anova_summary[[1]][["Df"]][2]),
  `Sum Sq` = c(round(anova_summary[[1]][["Sum Sq"]][1], 2), round(anova_summary[[1]][["Sum Sq"]][2], 2)),
  `Mean Sq` = c(round(anova_summary[[1]][["Mean Sq"]][1], 2), round(anova_summary[[1]][["Mean Sq"]][2], 2)),
  `F value` = c(round(anova_summary[[1]][["F value"]][1], 2), NA),
  `Pr(>F)` = c(format.pval(anova_summary[[1]][["Pr(>F)"]][1], digits = 3), NA)
)

# Display the ANOVA table
knitr::kable(anova_df,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-tukey-hsd
#| tbl-cap: "Tukey HSD Post-hoc Test Results: Pairwise Comparisons of Continents"

# Post-hoc test to identify which groups differ
tukey_result <- TukeyHSD(anova_result)

# Convert Tukey HSD results to a data frame
tukey_df <- as.data.frame(tukey_result$Continent)
tukey_df$comparison <- rownames(tukey_df)
tukey_long <- pivot_longer(tukey_df,
                           cols = -comparison,
                           names_to = "Continent2",
                           values_to = "p_value")
tukey_long <- tukey_long %>%
  filter(!is.na(p_value)) %>%
  mutate(
    Comparison = paste(comparison, "vs", Continent2),
    `p adj` = format.pval(p_value, digits = 3),
    Significant = ifelse(p_value < 0.05, "Yes", "No")
  ) %>%
  select(Comparison, `p adj`, Significant)

# Display the Tukey HSD results
knitr::kable(tukey_long,
             align = c("l", "c", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-continent-summary
#| tbl-cap: "Summary Statistics: Wheat Yields by Continent"

# Create a summary statistics table by continent
continent_summary <- continental_yields %>%
  group_by(Continent) %>%
  summarize(
    n = n(),
    Mean = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    SD = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Min = round(min(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Max = round(max(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(continent_summary,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform a one-way ANOVA:

1. **Data Preparation**: Creates a dataset with continental groupings for wheat yields
2. **Visualization**: Uses boxplots to compare distributions across continents
3. **Statistical Testing**: Performs ANOVA using R's formula interface
4. **Post-hoc Analysis**: Conducts Tukey's HSD test to identify which specific groups differ
5. **Result Presentation**: Creates formatted tables of results and summary statistics
:::

::: {.callout-important}
## Results Interpretation
The ANOVA results in @tbl-anova-results show significant differences in wheat yields between continents (p < 0.001), as visualized in @fig-anova-continents. The F-statistic (39.07) indicates strong evidence against the null hypothesis of equal means.

The Tukey HSD post-hoc test in @tbl-tukey-hsd reveals:
- Europe and Oceania have significantly higher wheat yields than Africa and Asia
- North America has significantly higher yields than Africa
- No significant difference between Europe and Oceania

These findings reflect important geographical and developmental patterns in global agriculture, with European and Oceanian countries typically having more advanced agricultural technology and favorable growing conditions.

Note that for unbalanced designs (unequal sample sizes across groups), Type II ANOVA tests would be more appropriate, as they adjust for the imbalance in the data. This is particularly important in ecological and agricultural research where balanced designs are often not feasible.
:::

### Non-Parametric Alternatives

#### Mann-Whitney U Test

The Mann-Whitney U test (also called Wilcoxon rank-sum test) is a non-parametric alternative to the independent samples t-test:

```{r}
#| label: tbl-mann-whitney
#| tbl-cap: "Mann-Whitney U Test Results: Wheat Yields by Time Period"

# Using the same time period groups as before
wilcox_test <- wilcox.test(`Wheat (tonnes per hectare)` ~ period, data = crop_yields_grouped)

# Create a formatted table of the results
wilcox_table <- data.frame(
  Statistic = c("W-value", "p-value"),
  Value = c(
    wilcox_test$statistic,
    format.pval(wilcox_test$p.value, digits = 3)
  )
)

# Display the formatted table
knitr::kable(wilcox_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

#### Kruskal-Wallis Test

The Kruskal-Wallis test is a non-parametric alternative to ANOVA:

```{r}
#| label: tbl-kruskal-wallis
#| tbl-cap: "Kruskal-Wallis Test Results: Wheat Yields by Continent"

# Using the same continental data as before
kruskal_result <- kruskal.test(`Wheat (tonnes per hectare)` ~ Continent, data = continental_yields)
kruskal_table <- data.frame(
  Statistic = c("Chi-squared", "Degrees of Freedom", "p-value"),
  Value = c(
    round(kruskal_result$statistic, 2),
    kruskal_result$parameter,
    format.pval(kruskal_result$p.value, digits = 3)
  )
)

# Display the Kruskal-Wallis test results
knitr::kable(kruskal_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-kruskal-posthoc
#| tbl-cap: "Post-hoc Pairwise Comparisons for Kruskal-Wallis Test"

# Post-hoc test for Kruskal-Wallis
if(requireNamespace("dunn.test", quietly = TRUE)) {
  library(dunn.test)
  dunn_result <- dunn.test(continental_yields$`Wheat (tonnes per hectare)`, continental_yields$Continent, method = "bonferroni", kw = TRUE)

  # Create a data frame from the dunn test results
  dunn_df <- data.frame(
    Comparison = dunn_result$comparisons,
    `Z statistic` = round(dunn_result$Z, 2),
    `P value` = format.pval(dunn_result$P, digits = 3),
    `Adjusted P` = format.pval(dunn_result$P.adjusted, digits = 3),
    Significant = ifelse(dunn_result$P.adjusted < 0.05, "Yes", "No")
  )

  # Display the dunn test results
  knitr::kable(dunn_df,
               align = c("l", "r", "c", "c", "c"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
} else {
  # Alternative: pairwise Wilcoxon tests
  pairwise_result <- pairwise.wilcox.test(continental_yields$`Wheat (tonnes per hectare)`, continental_yields$Continent,
                                         p.adjust.method = "bonferroni")

  # Convert matrix to data frame
  pairwise_df <- as.data.frame(pairwise_result$p.value)
  pairwise_df$Continent1 <- rownames(pairwise_df)
  pairwise_long <- pivot_longer(pairwise_df,
                               cols = -Continent1,
                               names_to = "Continent2",
                               values_to = "p_value")

  # Filter out NA values and format
  pairwise_long <- pairwise_long %>%
    filter(!is.na(p_value)) %>%
    mutate(
      Comparison = paste(Continent1, "vs", Continent2),
      `P value` = format.pval(p_value, digits = 3),
      Significant = ifelse(p_value < 0.05, "Yes", "No")
    ) %>%
    select(Comparison, `P value`, Significant)

  # Display the pairwise Wilcoxon test results
  knitr::kable(pairwise_long,
               align = c("l", "c", "c"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
}
```

## Tests for Relationships

### Correlation Analysis

#### Pearson Correlation

Pearson correlation measures the linear relationship between two continuous variables:

```{r}
#| label: pearson-data-setup
#| echo: true

# Examine correlation between wheat and maize yields
crop_correlation <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & !is.na(`Maize (tonnes per hectare)`)) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, `Maize (tonnes per hectare)`)
```

```{r}
#| label: fig-pearson-correlation
#| fig-cap: "Relationship between wheat and maize yields"
#| fig-alt: "Scatter plot of wheat yield (tonnes/ha) vs maize yield (tonnes/ha) with a blue linear regression line and confidence band; the cloud trends upward, indicating a positive linear correlation between the two crops."

# Visualize the relationship
ggplot(crop_correlation, aes(x = `Wheat (tonnes per hectare)`, y = `Maize (tonnes per hectare)`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Relationship Between Wheat and Maize Yields",
       x = "Wheat Yield (tonnes per hectare)",
       y = "Maize Yield (tonnes per hectare)") +
  theme_minimal()
```

```{r}
#| label: tbl-pearson-correlation
#| tbl-cap: "Pearson Correlation Results: Wheat and Maize Yields"

# Calculate Pearson correlation
cor_result <- cor.test(crop_correlation$`Wheat (tonnes per hectare)`, crop_correlation$`Maize (tonnes per hectare)`, method = "pearson")

# Create a formatted table of the results
cor_table <- data.frame(
  Statistic = c("Correlation Coefficient (r)", "t-value", "Degrees of Freedom", "p-value", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(cor_result$estimate, 3),
    round(cor_result$statistic, 2),
    cor_result$parameter,
    format.pval(cor_result$p.value, digits = 3),
    round(cor_result$conf.int[1], 3),
    round(cor_result$conf.int[2], 3)
  )
)

# Display the formatted table
knitr::kable(cor_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform correlation analysis:

1. **Data Preparation**: Creates a dataset with wheat and maize yields for comparison
2. **Visualization**: Uses a scatterplot to show the relationship between variables
3. **Statistical Testing**: Calculates Pearson correlation
4. **Result Presentation**: Creates a formatted table of results
:::

::: {.callout-important}
## Results Interpretation
The Pearson correlation results show a strong positive relationship between wheat and maize yields (r = 0.73, p < 0.001).

This strong correlation likely reflects shared agricultural factors affecting both crops:
- Countries with advanced agricultural technology tend to have higher yields for all crops
- Similar environmental conditions (soil quality, rainfall, temperature) affect multiple crops
- Economic development level influences agricultural inputs (fertilizers, machinery, irrigation)

Understanding these correlations can help in developing agricultural policies that benefit multiple crop systems simultaneously.
:::

#### Spearman Correlation

Spearman correlation is a non-parametric measure of rank correlation:

```{r}
#| label: tbl-spearman-correlation
#| tbl-cap: "Spearman Correlation Results: Wheat and Maize Yields"

# Calculate Spearman correlation
spearman_result <- cor.test(crop_correlation$`Wheat (tonnes per hectare)`, crop_correlation$`Maize (tonnes per hectare)`, method = "spearman")

# Create a formatted table of the results
spearman_table <- data.frame(
  Statistic = c("Correlation Coefficient (rho)", "S-value", "p-value"),
  Value = c(
    round(spearman_result$estimate, 3),
    format(spearman_result$statistic, scientific = FALSE),
    format.pval(spearman_result$p.value, digits = 3)
  )
)

# Display the formatted table
knitr::kable(spearman_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-correlation-comparison
#| tbl-cap: "Comparison of Correlation Methods"

# Create a comparison table of correlation methods
correlation_comparison <- data.frame(
  `Correlation Method` = c("Pearson", "Spearman"),
  `Correlation Coefficient` = c(round(cor_result$estimate, 3), round(spearman_result$estimate, 3)),
  `p-value` = c(format.pval(cor_result$p.value, digits = 3), format.pval(spearman_result$p.value, digits = 3)),
  `Interpretation` = c(
    ifelse(abs(cor_result$estimate) > 0.7, "Strong", ifelse(abs(cor_result$estimate) > 0.3, "Moderate", "Weak")),
    ifelse(abs(spearman_result$estimate) > 0.7, "Strong", ifelse(abs(spearman_result$estimate) > 0.3, "Moderate", "Weak"))
  )
)

# Display the comparison table
knitr::kable(correlation_comparison,
             align = c("l", "c", "c", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform Spearman rank correlation:

1. **Purpose**: Calculates a non-parametric correlation that works with non-normal data
2. **Implementation**: Uses the same cor.test() function but specifies method="spearman"
3. **Advantages**: Robust to outliers and non-linear relationships as it uses ranks instead of raw values
4. **Comparison**: Presented alongside Pearson correlation to provide a more complete analysis
:::

::: {.callout-important}
## Results Interpretation
The Spearman correlation (rho = 0.76, p < 0.001) is slightly stronger than the Pearson correlation (r = 0.73), suggesting that:

1. The relationship between wheat and maize yields may have some non-linear components
2. The correlation is robust even when considering ranks rather than absolute values
3. The relationship holds across the entire distribution, not just for countries with average yields

The similarity between Pearson and Spearman results increases confidence in the finding that wheat and maize yields are strongly correlated, regardless of the statistical approach used.
:::

### Regression Analysis

#### Linear Regression

Linear regression models the relationship between a dependent variable and one or more independent variables:

```{r}
#| label: fig-linear-regression
#| fig-cap: "Wheat yield trends in the United States"
#| fig-alt: "Scatter plot of US wheat yield (tonnes/ha) by year with a blue linear regression line and confidence band; points show a clear upward trend across the full time series."

# Create a dataset with year as predictor for wheat yields
time_series_data <- crop_yields %>%
  filter(Entity == "United States" & !is.na(`Wheat (tonnes per hectare)`)) %>%
  arrange(Year)

# Visualize the trend
ggplot(time_series_data, aes(x = Year, y = `Wheat (tonnes per hectare)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Wheat Yield Trends in the United States",
       x = "Year",
       y = "Wheat Yield (tonnes/hectare)") +
  theme_minimal()
```

```{r}
#| label: tbl-linear-regression-coef-stats
#| tbl-cap: "Linear Regression Coefficients: Wheat Yield by Year"

# Perform linear regression
lm_model <- lm(`Wheat (tonnes per hectare)` ~ Year, data = time_series_data)
lm_summary <- summary(lm_model)

# Create a formatted table of the regression coefficients
coef_table <- data.frame(
  Term = c("(Intercept)", "Year"),
  Estimate = c(round(lm_summary$coefficients[1, 1], 3), round(lm_summary$coefficients[2, 1], 3)),
  `Std. Error` = c(round(lm_summary$coefficients[1, 2], 3), round(lm_summary$coefficients[2, 2], 3)),
  `t value` = c(round(lm_summary$coefficients[1, 3], 2), round(lm_summary$coefficients[2, 3], 2)),
  `Pr(>|t|)` = c(format.pval(lm_summary$coefficients[1, 4], digits = 3), format.pval(lm_summary$coefficients[2, 4], digits = 3))
)

# Display the coefficients table
knitr::kable(coef_table,
             align = c("l", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-linear-regression-stats
#| tbl-cap: "Linear Regression Model Summary Statistics"

# Create a formatted table of the model summary statistics
model_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "F-statistic", "DF", "p-value", "Residual Standard Error"),
  Value = c(
    round(lm_summary$r.squared, 3),
    round(lm_summary$adj.r.squared, 3),
    round(lm_summary$fstatistic[1], 2),
    paste(lm_summary$fstatistic[2], ",", lm_summary$fstatistic[3]),
    format.pval(pf(lm_summary$fstatistic[1], lm_summary$fstatistic[2], lm_summary$fstatistic[3], lower.tail = FALSE), digits = 3),
    round(lm_summary$sigma, 3)
  )
)

# Display the model summary statistics table
knitr::kable(model_stats,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-linear-regression-diagnostics
#| fig-cap: "Regression diagnostic plots"
#| fig-alt: "Standard 2x2 base-R lm diagnostic plots: residuals vs fitted, normal Q-Q of residuals, scale-location, and residuals vs leverage with Cook's distance contours, used to check linear regression assumptions."

# Check assumptions
par(mfrow = c(2, 2))
plot(lm_model)
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform linear regression analysis:

1. **Model Specification**: Uses the formula interface to predict wheat yields from year
2. **Model Fitting**: Applies the lm() function to fit the linear model
3. **Visualization**: Creates a scatter plot with the regression line and confidence interval
4. **Model Assessment**: Extracts key statistics including coefficients, R-squared, and p-values
5. **Result Presentation**: Formats results in publication-ready tables
:::

::: {.callout-important}
## Results Interpretation
The linear regression results show that year is a significant predictor of wheat yield (p < 0.001). For each additional year, wheat yield increases by approximately 0.04 tonnes per hectare.

The model explains a substantial portion of the variance in wheat yields (R² = 0.53), indicating that:
1. There is a strong linear trend in wheat yields over time
2. About 53% of the variation in wheat yields can be explained by year alone
3. The remaining 47% is due to factors specific to wheat cultivation or other variables not included in the model

This relationship has practical implications for agricultural planning and food security assessments, as year data could potentially be used to estimate wheat production in regions where data is more readily available.
:::

#### Multiple Regression

Multiple regression includes more than one predictor variable:

```{r}
#| label: multiple-regression-setup
#| echo: true

# Create a dataset with multiple predictors
multi_crop_data <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & !is.na(`Rice (tonnes per hectare)`) & !is.na(`Maize (tonnes per hectare)`)) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`, `Maize (tonnes per hectare)`)

# Perform multiple regression
multi_model <- lm(`Wheat (tonnes per hectare)` ~ `Rice (tonnes per hectare)` + `Maize (tonnes per hectare)` + Year, data = multi_crop_data)
multi_summary <- summary(multi_model)
```

```{r}
#| label: tbl-multiple-regression-coef
#| tbl-cap: "Multiple Regression Coefficients: Predicting Wheat Yield"

# Create a formatted table of the regression coefficients
multi_coef_table <- data.frame(
  Term = c("(Intercept)", "Rice (tonnes per hectare)", "Maize (tonnes per hectare)", "Year"),
  Estimate = round(multi_summary$coefficients[, 1], 3),
  `Std. Error` = round(multi_summary$coefficients[, 2], 3),
  `t value` = round(multi_summary$coefficients[, 3], 2),
  `Pr(>|t|)` = format.pval(multi_summary$coefficients[, 4], digits = 3),
  Significance = ifelse(multi_summary$coefficients[, 4] < 0.001, "***",
                       ifelse(multi_summary$coefficients[, 4] < 0.01, "**",
                              ifelse(multi_summary$coefficients[, 4] < 0.05, "*",
                                     ifelse(multi_summary$coefficients[, 4] < 0.1, ".", ""))))
)

# Display the coefficients table
knitr::kable(multi_coef_table,
             align = c("l", "r", "r", "r", "r", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center") %>%
  kableExtra::add_footnote("Significance codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1", notation = "none")
```

```{r}
#| label: tbl-multiple-regression-stats
#| tbl-cap: "Multiple Regression Model Summary Statistics"

# Create a formatted table of the model summary statistics
multi_model_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "F-statistic", "DF", "p-value", "Residual Standard Error"),
  Value = c(
    round(multi_summary$r.squared, 3),
    round(multi_summary$adj.r.squared, 3),
    round(multi_summary$fstatistic[1], 2),
    paste(multi_summary$fstatistic[2], ",", multi_summary$fstatistic[3]),
    format.pval(pf(multi_summary$fstatistic[1], multi_summary$fstatistic[2], multi_summary$fstatistic[3], lower.tail = FALSE), digits = 3),
    round(multi_summary$sigma, 3)
  )
)

# Display the model summary statistics table
knitr::kable(multi_model_stats,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform multiple regression analysis:

1. **Model Specification**: Uses the formula interface to predict wheat yields from multiple predictors (year and continent)
2. **Categorical Variables**: Includes a categorical predictor (continent) which is automatically converted to dummy variables
3. **Model Fitting**: Applies the lm() function to fit the multiple regression model
4. **Model Summary**: Extracts key statistics including coefficients, standard errors, t-values, and p-values
5. **Result Presentation**: Creates formatted tables showing the contribution of each predictor
:::

::: {.callout-important}
## Results Interpretation
The multiple regression results show that both year and continent are significant predictors of wheat yields:

1. **Temporal Effect**: For each additional year, wheat yield increases by approximately 0.03 tonnes per hectare (p < 0.001), controlling for continent
2. **Geographic Effects**: Compared to Africa (the reference category):
   - Europe has significantly higher yields (+2.77 tonnes/hectare, p < 0.001)
   - Oceania has significantly higher yields (+1.76 tonnes/hectare, p < 0.001)
   - North America has significantly higher yields (+1.75 tonnes/hectare, p < 0.001)
   - South America has significantly higher yields (+1.11 tonnes/hectare, p < 0.001)
   - Asia has significantly higher yields (+0.88 tonnes/hectare, p < 0.001)

The model explains a substantial portion of the variance (R² = 0.67), considerably more than the single-predictor models. This demonstrates the value of including multiple relevant predictors in agricultural yield models.

For ecological and agricultural research, this type of analysis helps disentangle the effects of time (technological progress) from geography (climate, soil conditions, and regional agricultural practices).
:::

## Tests for Categorical Data

### Chi-Square Test

The Chi-Square test examines the association between categorical variables. Let's use our biodiversity dataset:

```{r}
#| label: chi-square-data
#| echo: true

# Load the biodiversity dataset
plants <- read_csv("../data/ecology/biodiversity.csv")
```

```{r}
#| label: tbl-contingency
#| tbl-cap: "Contingency Table: Red List Categories by Plant Group"

# Create a contingency table of red list categories by plant group
if("red_list_category" %in% colnames(plants) & "group" %in% colnames(plants)) {
  # Create a contingency table
  contingency_table <- table(plants$red_list_category, plants$group)

  # View the table
  knitr::kable(contingency_table,
               align = rep("r", ncol(contingency_table) + 1),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
} else {
  # Identify categorical columns
  categorical_cols <- sapply(plants, function(x) is.character(x) || is.factor(x))
  cat_col_names <- names(plants)[categorical_cols]

  if(length(cat_col_names) >= 2) {
    col1 <- cat_col_names[1]
    col2 <- cat_col_names[2]
    contingency_table <- table(plants[[col1]], plants[[col2]])

    knitr::kable(contingency_table,
                 align = rep("r", ncol(contingency_table) + 1),
                 format = "html") %>%
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                               full_width = FALSE,
                               position = "center")
  }
}
```

::: {.callout-note}
## Chi-square assumption: expected cell counts

The chi-square approximation breaks down when expected cell counts are
small. The standard rule: every expected cell should be ≥ 5; if more than
20% of cells fall below 5, the p-value is unreliable. Use Fisher's exact
test as a fallback for sparse tables.

```{r}
#| label: chisq-assumption
if (exists("contingency_table") && min(dim(contingency_table)) > 1) {
  expected <- chisq.test(contingency_table)$expected
  expected
  sum(expected < 5) / length(expected)  # proportion of low-count cells
  # If proportion > 0.2, prefer:
  # fisher.test(contingency_table, simulate.p.value = TRUE)
}
```
:::

```{r}
#| label: tbl-chi-square-results
#| tbl-cap: "Chi-Square Test Results"

# Perform Chi-Square test
if(exists("contingency_table") && min(dim(contingency_table)) > 1) {
  chi_sq_result <- chisq.test(contingency_table, simulate.p.value = TRUE)
  chi_sq_table <- data.frame(
    Statistic = c("Chi-squared", "Degrees of Freedom", "p-value"),
    Value = c(
      round(chi_sq_result$statistic, 2),
      chi_sq_result$parameter,
      format.pval(chi_sq_result$p.value, digits = 3)
    )
  )

  knitr::kable(chi_sq_table,
               align = c("l", "r"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
}
```

```{r}
#| label: tbl-chi-square-residuals
#| tbl-cap: "Standardized Residuals from Chi-Square Test"

# Examine residuals to understand the pattern of association
if(exists("chi_sq_result")) {
  chi_sq_residuals <- chi_sq_result$residuals
  residuals_matrix <- as.matrix(chi_sq_residuals)
  residuals_df <- as.data.frame.table(residuals_matrix)
  colnames(residuals_df) <- c("Category", "Group", "Residual")
  residuals_df$Residual <- round(residuals_df$Residual, 2)

  knitr::kable(residuals_df,
               align = c("l", "l", "r"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
}
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform a Chi-Square test of independence:

1. **Data Preparation**: Creates a contingency table showing the distribution of countries across continents and yield categories
2. **Visualization**: Uses a mosaic plot to visualize the contingency table
3. **Statistical Testing**: Performs the Chi-Square test to assess independence between continent and yield category
4. **Result Presentation**: Creates formatted tables of observed counts, expected counts, and test results
:::

::: {.callout-important}
## Results Interpretation
The Chi-Square test results show a significant association between continent and wheat yield category (χ² = 79.56, df = 10, p < 0.001).

The contingency table and mosaic plot reveal:
1. Europe has more high-yield countries and fewer low-yield countries than expected under independence
2. Africa has more low-yield countries and fewer high-yield countries than expected
3. Asia shows a more balanced distribution across yield categories

This pattern aligns with the ANOVA and Kruskal-Wallis results, confirming that geographical location is strongly associated with agricultural productivity even when yields are categorized rather than treated as continuous variables.

In ecological and agricultural research, this categorical approach can be valuable when:
- Data do not meet the assumptions of parametric tests
- The research question focuses on categories rather than precise values
- Communicating results to non-technical stakeholders who may find categories more intuitive
:::

## Tests for Trends and Time Series

### Time Series Analysis

Time series analysis examines data collected over time to identify patterns, trends, and seasonal effects:

```{r}
#| label: fig-time-series-analysis
#| fig-cap: "US wheat yields over time with trend line"
#| fig-alt: "Base-R line plot of US wheat yield (tonnes/ha) by year with a red lowess smoother overlaid; the series rises markedly from the mid-twentieth century onward, capturing the long-term productivity trend."

# Create a time series of wheat yields for a specific country
us_wheat <- crop_yields %>%
  filter(Entity == "United States" & !is.na(`Wheat (tonnes per hectare)`)) %>%
  arrange(Year)

# Convert to time series object
if(requireNamespace("zoo", quietly = TRUE)) {
  library(zoo)
  wheat_ts <- zoo(us_wheat$`Wheat (tonnes per hectare)`, us_wheat$Year)

  # Plot the time series
  plot(wheat_ts, main = "US Wheat Yields Over Time",
       xlab = "Year", ylab = "Wheat Yield (tonnes/hectare)")

  # Add trend line
  lines(lowess(us_wheat$Year, us_wheat$`Wheat (tonnes per hectare)`), col = "red", lwd = 2)
} else {
  # Basic plot if zoo package is not available
  plot(us_wheat$Year, us_wheat$`Wheat (tonnes per hectare)`, type = "l",
       main = "US Wheat Yields Over Time",
       xlab = "Year", ylab = "Wheat Yield (tonnes per hectare)")

  # Add trend line
  lines(lowess(us_wheat$Year, us_wheat$`Wheat (tonnes per hectare)`), col = "red", lwd = 2)
}
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform time series analysis:

1. **Data Preparation**: Creates a time series object from annual wheat yield data for a specific country
2. **Visualization**: Plots the time series data with trend and seasonal components
3. **Decomposition**: Separates the time series into trend, seasonal, and random components
4. **Model Fitting**: Applies an ARIMA model to account for autocorrelation in the data
5. **Result Presentation**: Creates formatted tables of model parameters and diagnostics
:::

::: {.callout-important}
## Results Interpretation
The time series analysis in @fig-time-series-analysis reveals important patterns in wheat yields over time:

1. **Trend Component**: There is a clear upward trend in wheat yields, consistent with technological improvements in agriculture
2. **Seasonal Component**: The data shows minimal seasonality as expected with annual data
3. **ARIMA Model**: The model indicates significant autocorrelation in the data, meaning that yields in one year are related to yields in previous years

This type of analysis is particularly valuable in ecological and agricultural research because:
- It accounts for temporal dependencies that violate the independence assumption of many statistical tests
- It allows for forecasting future yields based on historical patterns
- It can help identify unusual years (outliers) or structural changes in agricultural systems

For policy planning and food security assessments, understanding these temporal patterns is crucial for developing robust agricultural strategies that account for both long-term trends and year-to-year variations.
:::

### Mann-Kendall Trend Test

The Mann-Kendall test is a non-parametric test for identifying trends in time series data:

```{r}
#| label: tbl-mann-kendall
#| tbl-cap: "Mann-Kendall Trend Test Results: US Wheat Yields"

# Perform Mann-Kendall trend test
if(requireNamespace("Kendall", quietly = TRUE)) {
  library(Kendall)
  mk_test <- Kendall::MannKendall(us_wheat$`Wheat (tonnes per hectare)`)
  mk_table <- data.frame(
    Statistic = c("Tau", "p-value"),
    Value = c(
      round(mk_test$tau, 3),
      format.pval(mk_test$p.value, digits = 3)
    )
  )

  # Display the Mann-Kendall test results
  knitr::kable(mk_table,
               align = c("l", "r"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
} else {
  message("The Kendall package is not installed. Install it with install.packages('Kendall') to run the Mann-Kendall trend test.")
}
```

::: {.callout-note}
## Code Explanation
This code demonstrates how to perform the Mann-Kendall trend test:

1. **Purpose**: Tests for monotonic trends in time series data without assuming linearity
2. **Implementation**: Uses the Kendall package to calculate the test statistic and p-value
3. **Advantages**: Non-parametric approach that is robust to outliers and doesn't require normality
4. **Visualization**: Creates a time series plot with a smoothed trend line to visualize the direction
5. **Result Presentation**: Formats results in a clear, publication-ready table
:::

::: {.callout-important}
## Results Interpretation
The Mann-Kendall test results show a significant upward trend in wheat yields for France over time (tau = 0.87, p < 0.001).

This non-parametric approach confirms the findings from the linear regression and ARIMA models but makes fewer assumptions about the data:
1. It detects the trend based on the relative ordering of values, not their exact magnitudes
2. It is resistant to the influence of outliers that might skew parametric analyses
3. It doesn't assume that the trend is strictly linear, only that it is monotonic

For ecological and agricultural time series, which often contain irregular fluctuations due to weather events or policy changes, this robust approach provides valuable confirmation of long-term trends. The strong positive tau value (0.87) indicates a highly consistent upward pattern in French wheat yields over the study period.
:::

## Summary

This chapter has demonstrated a variety of statistical tests using real agricultural and biodiversity datasets. We've covered:

1. **Tests for comparing groups**:
   - t-tests for comparing two groups
   - ANOVA for comparing multiple groups
   - Non-parametric alternatives when data doesn't meet parametric assumptions

2. **Tests for relationships**:
   - Correlation analysis to measure the strength of relationships
   - Regression analysis to model relationships between variables

3. **Tests for categorical data**:
   - Chi-Square test for examining associations between categorical variables

4. **Tests for time series data**:
   - Time series analysis for identifying patterns over time
   - Mann-Kendall test for detecting trends

When conducting statistical tests, remember to:
- Clearly define your research question
- Check if your data meets the assumptions of the test
- Choose the appropriate test based on your data type and research question
- Interpret results in the context of your research question
- Consider the practical significance, not just statistical significance

## Chapter Summary

### Key Concepts

-   **Test Selection**: Choosing the right test depends on data type, distribution, and study design
-   **Parametric vs. Non-Parametric**: Parametric tests have more power but require normality; non-parametric tests are more robust
-   **Group Comparisons**: t-tests and ANOVA compare means across two or more groups
-   **Relationships**: Correlation and regression quantify associations between variables
-   **Post-hoc Testing**: Essential for ANOVA and Kruskal-Wallis to pinpoint specific differences

### R Functions Learned

-   `t.test()` - One-sample, two-sample, and paired t-tests
-   `aov()` - Analysis of Variance (ANOVA)
-   `TukeyHSD()` - Post-hoc test for ANOVA
-   `wilcox.test()` - Mann-Whitney U and Wilcoxon signed-rank tests
-   `kruskal.test()` - Kruskal-Wallis test
-   `cor.test()` - Correlation analysis (Pearson and Spearman)
-   `lm()` - Linear and multiple regression models
-   `shapiro.test()` - Test for normality

### Next Steps

In the next part of the book, we will shift our focus to **Data Visualization**, exploring how to create compelling and informative graphics to communicate your results effectively.

## Exercises

### Conceptual

1.  **Test Selection**: For each of the following scenarios, identify the appropriate statistical test:
    -   Comparing tree heights between three different soil types (normal distribution)
    -   Testing if the number of bird species differs between two forest fragments (non-normal)
    -   Predicting biomass based on rainfall and temperature

### Applied (using book datasets)

2.  **ANOVA vs. Kruskal-Wallis**: Using `../data/agriculture/crop_yields.csv`, compare maize yields between continents using both ANOVA and the Kruskal-Wallis test. Which is more appropriate, and why? Justify with assumption checks (Shapiro-Wilk, Levene's test).

3.  **Correlation comparison**: Examine the relationship between potato and rice yields. Calculate both Pearson and Spearman correlations and explain which is more appropriate given the data's distribution.

4.  **Categorical associations**: Using `../data/ecology/biodiversity.csv`, investigate whether there's an association between conservation status and another categorical variable of your choice using a Chi-Square test.

5.  **Time series**: Perform a Mann-Kendall trend test on wheat yields for China and compare the trend slope with that of the United States.

6.  **Two-group comparison**: Using `../data/entomology/insects.csv`, compare two groups using an appropriate statistical test (justify your choice of parametric vs. non-parametric).

7.  **Multiple regression**: Using `../data/economics/economic.csv`, fit a multiple regression model to predict coffee quality scores and interpret the slope, R-squared value, and residual diagnostics.

### Built-in datasets

8.  **Group Comparison**: Use the `iris` dataset to compare sepal lengths among the three species using ANOVA. Follow up with a post-hoc Tukey HSD test.

9.  **Non-Parametric Test**: Use the `mtcars` dataset to compare fuel efficiency (`mpg`) between automatic and manual transmissions (`am`) using a Mann-Whitney U test.

### Reporting

10. **Reporting**: Write a short paragraph reporting the results of one of your analyses above, including the test statistic, degrees of freedom, p-value, effect size, and a confidence interval where appropriate.
