---
title: "Solutions — 01-introduction"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Install and explore

No code answer. Mark for: students naming the four RStudio panes (Source, Console, Environment/History, Files/Plots/Help) and locating the project root.

## Exercise 2 — Load the tidyverse

```{r}
library(tidyverse)
```

Loaded packages: ggplot2, tibble, tidyr, readr, purrr, dplyr, stringr, forcats, lubridate. Watch for masking warnings (`filter` and `lag` from `stats`) — they should know what got masked.

## Exercise 3 — Explore iris

```{r}
head(iris)
glimpse(iris)
summary(iris)
```

**Interpretation.** `head()` prints rows, `glimpse()` shows types and a few values per column, `summary()` gives per-column statistics. Each one answers a different question; good students use all three rather than picking a favourite.

## Exercise 4 — Pipe rewrite

```{r}
c(4, 9, 16, 25, 36) %>% sqrt() %>% mean() %>% round(2)
```

**Interpretation.** Reads left-to-right and matches the order of operations. The nested version reads outside-in, which is why long pipelines beat deeply nested calls.

## Exercise 5 — Grouped summary

```{r}
iris %>%
  group_by(Species) %>%
  summarise(across(Sepal.Length, list(mean = mean, sd = sd)))
```

**Interpretation.** `across()` applies the same functions to multiple columns without copy-pasting. Equivalent to `summarise(mean = mean(Sepal.Length), sd = sd(Sepal.Length))` for one column.

## Exercise 6 — Scatter plot

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, colour = Species)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Sepal dimensions by iris species")
```

**Interpretation.** *setosa* clearly separates; *versicolor* and *virginica* overlap. Good students mention the cluster structure before reading off any numeric trend.

## Exercise 7 — Research question

No code answer. Expect a question of the form *does X affect Y in Z system*, a data type (counts, measurements, occurrences), and a visualisation choice that matches the variable types.

## Common student pitfalls

::: {.callout-tip}
- Loading `tidyverse` inside every code chunk.
- Using `attach()` to avoid typing the data frame name.
- Reading `summary()` output as if it's a test result.
- Reaching for base-R `aggregate()` when `group_by() + summarise()` is cleaner.
:::
