---
title: "Solutions — 02-data-basics"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(palmerpenguins)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.
>
> Note: the chapter points to `data/environmental/climate_data.csv` for
> penguins, which is a known dataset/filename mismatch
> (see `data/MISMATCHES.md`). Solutions below use the `palmerpenguins`
> package directly so they run reliably.

## Exercise 1 — Counts by species and island

```{r}
penguins %>%
  count(species, island) %>%
  pivot_wider(names_from = island, values_from = n, values_fill = 0)
```

**Interpretation.** Wide format is easier to scan when the question is "which species is on which island?". Long format (the raw `count()` output) is easier to plot.

## Exercise 2 — Mean and SD per species

```{r}
penguins %>%
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g) %>%
  group_by(species) %>%
  summarise(across(c(bill_length_mm, bill_depth_mm, body_mass_g),
                   list(mean = mean, sd = sd), .names = "{.col}_{.fn}"))
```

**Interpretation.** Adelie has shortest bills and lightest body mass; Gentoo is heaviest. Watch for students who forget `drop_na()` and report NaN means.

## Exercise 3 — Flipper-to-mass ratio

```{r}
penguins %>%
  drop_na(flipper_length_mm, body_mass_g) %>%
  mutate(flip_per_g = flipper_length_mm / body_mass_g) %>%
  group_by(species) %>%
  summarise(mean_ratio = mean(flip_per_g))
```

**Interpretation.** A higher flipper-length-per-gram ratio implies relatively longer flippers for body size — a possible swimming-efficiency proxy. Caution students against over-interpreting a ratio without considering body shape.

## Exercise 4 — Bill length vs bill depth

```{r}
ggplot(penguins %>% drop_na(),
       aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "Bill morphology by species",
       x = "Bill length (mm)", y = "Bill depth (mm)")
```

**Interpretation.** This is a classic Simpson's paradox illustration — overall the two variables look uncorrelated, but within each species they're positively related. Worth flagging in class.

## Exercise 5 — Wheat trends for one country

```{r}
crops <- read_csv("../data/agriculture/crop_yields.csv")
crops %>%
  filter(country == "Australia") %>%
  ggplot(aes(year, wheat_yield)) +
  geom_line() + geom_smooth(method = "loess") +
  theme_minimal() +
  labs(title = "Wheat yield in Australia over time")
```

**Interpretation.** LOESS reveals departures from linearity better than a straight line. A flat line with high variance is just as interesting as a clear trend.

## Exercise 6 — Mass by sex

```{r}
penguins %>%
  drop_na(sex, body_mass_g) %>%
  ggplot(aes(sex, body_mass_g, fill = sex)) +
  geom_violin(alpha = 0.4) +
  geom_boxplot(width = 0.2, show.legend = FALSE) +
  facet_wrap(~species) +
  theme_minimal()
```

**Interpretation.** Males are heavier than females across all three species — sexual size dimorphism. Faceting by species prevents the between-species variation from swamping the within-species comparison.

## Common student pitfalls

::: {.callout-tip}
- Reading the wrong file because chapter text and filenames disagree.
- Summarising without dropping NAs and then reporting NaN.
- Pooling species when comparing sexes (Simpson's paradox).
- Plotting wheat yield without sorting by year first.
:::
