---
title: "Solutions — 03-exploratory-analysis"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(viridis)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Biodiversity EDA

```{r}
bio <- read_csv("../data/ecology/biodiversity.csv")
glimpse(bio)
summary(bio)

# Univariate distributions for every numeric column
bio %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  ggplot(aes(val)) +
  geom_histogram(bins = 30, fill = viridis(1)) +
  facet_wrap(~ var, scales = "free") +
  theme_minimal()
```

**Interpretation.** Look for skew, multi-modality, and out-of-range values.
Flag columns where median << mean (right skew) — these are candidates for
log-transform before downstream tests.

## Exercise 2 — Distribution of a second crop

```{r}
crops <- read_csv("../data/agriculture/crop_yields.csv")
# Replace `rice_yield` with another crop column present in the file.
ggplot(crops, aes(x = rice_yield)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = viridis(1)) +
  geom_density(color = "red") +
  theme_minimal()
```

**Interpretation.** Compare central tendency, spread, and skew to wheat.
Rice yields typically have a heavier right tail than wheat in this dataset.

## Exercise 3 — Yield vs latitude

Requires joining a geography table to the crops table. If the bundled data
lacks latitude, students should merge against a country-centroid lookup
(e.g., `rnaturalearth::ne_countries()`). Mark down for any solution that
silently drops countries with missing matches without noting the drop.

## Exercise 4 — Largest improvement over time

```{r}
crops %>%
  group_by(country) %>%
  filter(n() >= 2) %>%
  summarise(
    delta = last(wheat_yield) - first(wheat_yield),
    rel = delta / first(wheat_yield),
    .groups = "drop"
  ) %>%
  arrange(desc(rel)) %>%
  slice_head(n = 10)
```

**Interpretation.** Relative change (`rel`) is the right metric — absolute
deltas reward countries that already had high yields.

## Exercise 5 — Wheat-to-rice ratio

```{r}
crops %>%
  filter(country %in% c("China", "United States", "India")) %>%
  mutate(ratio = wheat_yield / rice_yield) %>%
  ggplot(aes(year, ratio, color = country)) +
  geom_line() +
  scale_color_viridis_d() +
  theme_minimal()
```

## Exercise 6 — Spatial EDA

Same pipeline as Exercise 1 applied to `../data/geography/spatial.csv`.

## Common student pitfalls

::: {.callout-tip}
- Reading from `docs/data/` instead of `../data/` — old path, now broken.
- Not setting `set.seed()` before any sampling step.
- Reporting only the mean for skewed columns; require median + IQR too.
:::
