---
title: "Solutions — 04-hypothesis-testing"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Two-sample t-test on forest inventory

```{r}
forest <- read_csv("../data/forestry/forest_inventory.csv")

# Replace `species` and the level names with what's actually in the file.
two_species <- forest %>% filter(species %in% c("Pinus", "Eucalyptus"))

# Assumption checks
by_sp <- split(two_species$dbh, two_species$species)
shapiro.test(by_sp[[1]])
shapiro.test(by_sp[[2]])
var.test(dbh ~ species, data = two_species)

# Welch's t-test (default — robust to unequal variance)
t.test(dbh ~ species, data = two_species)
```

**Interpretation.** Report t, df, p, and a Cohen's d. If Shapiro fails on
either group, switch to Wilcoxon rank-sum and note the change.

## Exercise 2 — One-sample t-test against reference

```{r}
bio <- read_csv("../data/ecology/biodiversity.csv")
t.test(bio$species_richness, mu = 15)
```

**Interpretation.** H0: mean richness = 15. Reject if p < 0.05 and the CI
excludes 15. Always state direction (above/below 15) — students often omit
this.

## Exercise 3 — Paired t-test on crop yields

```{r}
crops <- read_csv("../data/agriculture/crop_yields.csv")
paired <- crops %>%
  group_by(country) %>%
  summarise(wheat = mean(wheat_yield, na.rm = TRUE),
            rice = mean(rice_yield, na.rm = TRUE),
            .groups = "drop") %>%
  drop_na()

# Assumption: differences are approximately normal
shapiro.test(paired$wheat - paired$rice)

t.test(paired$wheat, paired$rice, paired = TRUE)
```

**Interpretation.** Paired design controls for country-level confounders
(climate, infrastructure). If shapiro fails, use `wilcox.test(..., paired = TRUE)`.

## Exercise 4 — Normality checks

```{r}
x <- bio$species_richness
qqnorm(x); qqline(x, col = "red")
shapiro.test(x)
```

**Interpretation.** Q-Q deviations at the tails indicate heavy tails;
S-shape indicates skew. Shapiro is sensitive at large n — at n > 5000 it
will reject even tiny deviations, so weight the visual check higher.

## Exercise 5 — 95% CI

```{r}
clim <- read_csv("../data/environmental/climate_data.csv")
ci <- t.test(clim$temperature)$conf.int
ci
```

**Interpretation.** "We are 95% confident the long-run mean temperature
lies in [lower, upper] °C." Do **not** say "there is a 95% chance the
true mean is in this interval" — that's a Bayesian statement.

## Exercise 6 — Design your own test

Marking rubric:
- States H0 and H1 in symbolic form (e.g., μ1 = μ2 vs μ1 ≠ μ2).
- Identifies one-sided vs two-sided and justifies.
- Names the test and its assumptions.
- States α and how multiple comparisons (if any) will be handled.
- Specifies effect size to report alongside p.

## Common student pitfalls

::: {.callout-tip}
- Treating p < 0.05 as proof of H1. Remind them: it's evidence against H0
  given the assumptions hold.
- Forgetting to check normality before parametric tests.
- Using two-sample t-test on paired data — inflates variance, lowers power.
- Quoting "p = 0.000" — should be "p < 0.001".
:::
