---
title: "Solutions — 05-statistical-tests"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(car)        # Levene
library(broom)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Test selection (conceptual)

- Tree heights across 3 soil types, normal: **one-way ANOVA**, post-hoc Tukey HSD.
- Bird species across 2 fragments, non-normal: **Mann-Whitney U** (`wilcox.test`).
- Predicting biomass from rainfall + temperature: **multiple linear regression**.

## Exercise 2 — ANOVA vs Kruskal-Wallis on maize

```{r}
crops <- read_csv("../data/agriculture/crop_yields.csv")

# Assumption checks
shapiro.test(crops$maize_yield[crops$continent == "Africa"])
car::leveneTest(maize_yield ~ continent, data = crops)

aov_fit <- aov(maize_yield ~ continent, data = crops)
summary(aov_fit)

kruskal.test(maize_yield ~ continent, data = crops)
```

**Interpretation.** If either assumption fails, Kruskal-Wallis is the
correct test — it ranks the data and so doesn't require normality or
equal variance. ANOVA is more powerful when assumptions hold.

## Exercise 3 — Pearson vs Spearman

```{r}
with(crops, cor.test(potato_yield, rice_yield, method = "pearson"))
with(crops, cor.test(potato_yield, rice_yield, method = "spearman"))
```

**Interpretation.** Pearson assumes linear association and normality;
Spearman ranks the data and captures any monotonic relationship. If the
scatterplot shows curvature or heavy tails, prefer Spearman.

## Exercise 4 — Chi-square on biodiversity

```{r}
bio <- read_csv("../data/ecology/biodiversity.csv")
tab <- table(bio$conservation_status, bio$habitat_type)
chi <- chisq.test(tab)
chi

# Check expected cell counts
chi$expected
```

**Interpretation.** If any expected cell < 5, fall back to
`fisher.test(tab)`. Report χ², df, p, and Cramer's V for effect size.

## Exercise 5 — Mann-Kendall trend

```{r}
library(Kendall)
china <- crops %>% filter(country == "China") %>% arrange(year)
us    <- crops %>% filter(country == "United States") %>% arrange(year)

MannKendall(china$wheat_yield)
MannKendall(us$wheat_yield)

# Sen's slope for the rate of change
library(trend)
sens.slope(china$wheat_yield)
sens.slope(us$wheat_yield)
```

**Interpretation.** Mann-Kendall tests for monotonic trend; Sen's slope
gives the median rate of change per year. Compare slopes directly.

## Exercise 6 — Two-group comparison on insects

```{r}
insects <- read_csv("../data/entomology/insects.csv")
# Choose two groups present in the data; verify with table() first.
shapiro.test(insects$count[insects$group == "A"])
# If normal + equal var → t.test; else → wilcox.test
wilcox.test(count ~ group, data = insects %>% filter(group %in% c("A", "B")))
```

## Exercise 7 — Multiple regression on coffee quality

```{r}
econ <- read_csv("../data/economics/economic.csv")
fit <- lm(quality_score ~ price + altitude + rainfall, data = econ)
summary(fit)

# Diagnostics
par(mfrow = c(2, 2)); plot(fit); par(mfrow = c(1, 1))
```

**Interpretation.** Each slope is the change in quality per unit predictor
**holding the others constant**. R² is the proportion of variance
explained. Residual plots should show no pattern; QQ should be linear.

## Exercise 8 — iris ANOVA + Tukey

```{r}
fit_iris <- aov(Sepal.Length ~ Species, data = iris)
summary(fit_iris)
TukeyHSD(fit_iris)
```

**Interpretation.** All three pairwise differences are significant.
*setosa* has the shortest sepals; *virginica* the longest.

## Exercise 9 — mtcars Mann-Whitney

```{r}
wilcox.test(mpg ~ am, data = mtcars)
```

**Interpretation.** Manual transmissions show higher mpg in this sample.
Effect size via rank-biserial correlation if needed.

## Exercise 10 — Reporting paragraph (template)

> A one-way ANOVA on `Sepal.Length` across the three *iris* species was
> statistically significant, *F*(2, 147) = 119.3, *p* < 0.001, η² = 0.62.
> Tukey HSD revealed all pairwise comparisons significant at α = 0.05;
> *virginica* (M = 6.59, 95% CI [6.45, 6.72]) was longest, followed by
> *versicolor* (M = 5.94, [5.81, 6.07]) and *setosa* (M = 5.01, [4.91, 5.10]).

## Common student pitfalls

::: {.callout-tip}
- Running ANOVA without checking Levene's test; using Tukey when groups
  are unequal in variance.
- Reporting p without effect size.
- Confusing statistical significance with practical importance.
- Using Pearson on non-linear or heavy-tailed data.
- Forgetting to report degrees of freedom.
:::
