---
title: "Solutions — 06-visualization"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(viridis)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Distribution of plant species across taxonomic groups

```{r}
bio <- read_csv("../data/ecology/biodiversity.csv")

bio %>%
  count(taxonomic_group, sort = TRUE) %>%
  ggplot(aes(x = fct_reorder(taxonomic_group, n), y = n, fill = taxonomic_group)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(x = NULL, y = "Number of species",
       title = "Species count by taxonomic group") +
  theme_minimal()
```

**Interpretation.** A horizontal ordered bar chart is the right default here because group labels are long and counts span a wide range. Flag students who use a pie chart or who leave the bars in alphabetical order.

## Exercise 2 — Rice yield trends for top 5 producers

```{r}
crops <- read_csv("../data/agriculture/crop_yields.csv")

top5 <- crops %>%
  group_by(country) %>%
  summarise(total = sum(rice_yield, na.rm = TRUE)) %>%
  slice_max(total, n = 5) %>%
  pull(country)

crops %>%
  filter(country %in% top5) %>%
  ggplot(aes(x = year, y = rice_yield, colour = country)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d() +
  labs(x = "Year", y = "Rice yield (t/ha)",
       title = "Rice yields, top 5 producers") +
  theme_minimal()
```

**Interpretation.** Lines are appropriate for ordered time data. Watch for students who use bars or who forget to filter before plotting (40+ overlapping lines is unreadable).

## Exercise 3 — Pairs plot of spatial numeric variables

```{r}
spatial <- read_csv("../data/geography/spatial.csv")

spatial %>%
  select(where(is.numeric)) %>%
  select(1:5) %>%       # keep it readable
  GGally::ggpairs()
```

**Interpretation.** `GGally::ggpairs()` gives scatter, density, and correlation in one grid. Caution students against pairing more than ~6 variables — the grid becomes uninterpretable.

## Exercise 4 — Conservation status across habitat types

```{r}
bio %>%
  count(habitat_type, conservation_status) %>%
  ggplot(aes(x = habitat_type, y = n, fill = conservation_status)) +
  geom_col(position = "fill") +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Habitat", y = "Proportion",
       fill = "Status",
       title = "Conservation status composition by habitat") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

**Interpretation.** A 100% stacked bar reveals composition; raw counts hide differences when habitat totals vary. A grouped (dodged) bar is acceptable if the student wants to show absolute counts too.

## Exercise 5 — Coffee correlation heatmap

```{r}
econ <- read_csv("../data/economics/economic.csv")

econ %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r") %>%
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(-1, 1), option = "C") +
  geom_text(aes(label = round(r, 2)), size = 3) +
  labs(title = "Correlation matrix, coffee economics") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank())
```

**Interpretation.** Diverging palette only works when zero is meaningful — here r = 0 means no linear association, so a perceptually uniform scale like viridis option C is fine. Annotating cells with the rounded coefficient is good practice for small matrices.

## Exercise 6 — Animated yield trend (gganimate)

```{r, eval=FALSE}
library(gganimate)

one_country <- crops %>% filter(country == "China")

p <- ggplot(one_country, aes(x = year, y = rice_yield)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3, colour = "#1f78b4") +
  labs(title = "Rice yield in China — year: {frame_along}",
       x = "Year", y = "Yield (t/ha)") +
  theme_minimal() +
  transition_reveal(year)

animate(p, nframes = 80, fps = 10)
```

**Interpretation.** `transition_reveal` is the right choice for cumulative time-series animation. Mark down students who animate without a clear pedagogic reason — animation should reveal something a static plot cannot.

## Common student pitfalls

::: {.callout-tip}
- Pie charts for more than three categories.
- Default `ggplot` grey theme in a publication figure.
- Missing axis units, missing titles, illegible angled labels.
- Using a rainbow palette instead of a colour-blind-safe one.
- Plotting 30+ countries without filtering — turns a line plot into spaghetti.
- Forgetting `alt` text or a figure caption when the chapter explicitly asked for accessibility.
:::
