---
title: "Solutions — 07-advanced-visualization"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(patchwork)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Publication-ready theme customization

```{r}
penguins <- palmerpenguins::penguins %>% drop_na(bill_length_mm, body_mass_g, species)

ggplot(penguins, aes(bill_length_mm, body_mass_g, colour = species)) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  labs(x = "Bill length (mm)", y = "Body mass (g)",
       colour = "Species",
       title = "Bill length predicts body mass",
       subtitle = "Palmer penguins, Antarctica",
       caption = "Data: palmerpenguins") +
  theme_minimal(base_size = 12, base_family = "sans") +
  theme(plot.title = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey40"),
        panel.grid.minor = element_blank(),
        legend.position = "top")
```

**Interpretation.** The accepted answer should pick a colour-blind-safe palette, set an explicit base size, drop chartjunk (minor gridlines), and add a caption with the data source. Default `theme_grey()` is not publication-ready.

## Exercise 2 — Three-panel figure with patchwork

```{r, fig.width=10, fig.height=4}
p1 <- ggplot(penguins, aes(bill_length_mm, body_mass_g, colour = species)) +
  geom_point() + theme_minimal() + labs(title = "Scatter")

p2 <- ggplot(penguins, aes(species, body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE) + theme_minimal() + labs(title = "Box")

p3 <- penguins %>% count(species) %>%
  ggplot(aes(species, n, fill = species)) +
  geom_col(show.legend = FALSE) + theme_minimal() + labs(title = "Bar")

(p1 | p2 | p3) +
  plot_annotation(title = "Penguin morphology — three views",
                  tag_levels = "A")
```

**Interpretation.** `patchwork` operators `|` (side-by-side) and `/` (stacked) compose plots without manual `grid.arrange`. Tag levels A/B/C are expected for any multi-panel figure submitted to a journal.

## Exercise 3 — Interactive plotly conversion

```{r, eval=FALSE}
library(plotly)

p <- ggplot(penguins,
            aes(bill_length_mm, body_mass_g,
                colour = species,
                text = paste("Island:", island,
                             "<br>Sex:", sex))) +
  geom_point() + theme_minimal()

ggplotly(p, tooltip = "text")
```

**Interpretation.** Custom `text` aesthetic plus `tooltip = "text"` gives readable hover info. Mark down submissions that leave the default tooltip showing internal column names.

## Exercise 4 — Site distribution map

```{r, eval=FALSE}
library(sf)
library(rnaturalearth)

sites <- tibble(
  site = c("Lae", "Madang", "Wewak"),
  lon  = c(146.99, 145.78, 143.63),
  lat  = c(-6.73, -5.22, -3.55)
)

png <- ne_countries(country = "Papua New Guinea", returnclass = "sf")

ggplot() +
  geom_sf(data = png, fill = "grey95", colour = "grey60") +
  geom_point(data = sites, aes(lon, lat), colour = "#d95f02", size = 3) +
  ggrepel::geom_text_repel(data = sites, aes(lon, lat, label = site)) +
  labs(title = "Study sites in Papua New Guinea") +
  theme_minimal()
```

**Interpretation.** A proper map needs a base layer, a projection that does not distort the region of interest, and labelled points. Plotting raw lat/lon as a scatter without a coastline is not a map.

## Exercise 5 — Time-series animation

```{r, eval=FALSE}
library(gganimate)
crops <- read_csv("../data/agriculture/crop_yields.csv")

p <- crops %>%
  filter(country %in% c("China", "India", "United States")) %>%
  ggplot(aes(year, rice_yield, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(title = "Rice yield, year: {frame_along}",
       x = "Year", y = "t/ha") +
  theme_minimal() +
  transition_reveal(year)

animate(p, nframes = 100, fps = 12, width = 700, height = 400)
```

**Interpretation.** Animation justifies itself only when motion encodes meaningful change. The label `{frame_along}` ties the title to the current frame so viewers always see which year is being drawn.

## Common student pitfalls

::: {.callout-tip}
- Reaching for animation when a small-multiples facet would do the job.
- Plotly tooltips dumping raw column names.
- Maps without a base layer or without a sensible projection.
- Patchwork figures with mismatched themes between panels — readers should not need to relearn the chart style three times.
- Overusing colour as decoration rather than encoding.
:::
