---
title: "Solutions — 08-regression"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.
>
> The chapter has 15 exercises across Base R, tidymodels, and Advanced
> tiers. Worked solutions for a representative subset are below; the
> remainder follow the same patterns.

## Exercise 1 — Body mass from bill length + bill depth

```{r}
peng <- penguins %>% drop_na(body_mass_g, bill_length_mm, bill_depth_mm)
fit <- lm(body_mass_g ~ bill_length_mm + bill_depth_mm, data = peng)
summary(fit)
par(mfrow = c(2, 2)); plot(fit); par(mfrow = c(1, 1))
```

**Interpretation.** Both slopes are significant and positive once species is ignored, but Simpson's paradox lurks — adding `species` reverses the bill-depth coefficient. Good students should notice the residuals-vs-fitted clustering and refit with species included.

## Exercise 2 — Interaction terms

```{r}
fit_main <- lm(body_mass_g ~ bill_length_mm + bill_depth_mm + species, data = peng)
fit_int  <- lm(body_mass_g ~ bill_length_mm * species + bill_depth_mm, data = peng)
anova(fit_main, fit_int)
AIC(fit_main, fit_int)
```

**Interpretation.** An interaction is justified if the ANOVA p < 0.05 *and* AIC drops by ≥ 2. Don't add interactions just because they're available.

## Exercise 3 — AIC model selection

```{r}
full <- lm(body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species + sex,
           data = peng %>% drop_na())
step(full, direction = "both", trace = 0) %>% summary()
```

**Interpretation.** `step()` is convenient for teaching but penalises complexity gently; for real research, prefer `MuMIn::dredge()` with explicit candidate sets.

## Exercise 4 — Multinomial logistic regression

```{r}
library(nnet)
mfit <- multinom(species ~ bill_length_mm + bill_depth_mm + flipper_length_mm,
                 data = peng, trace = FALSE)
pred <- predict(mfit, peng)
mean(pred == peng$species)
table(pred, peng$species)
```

**Interpretation.** Accuracy should exceed 95%. Watch for students reporting accuracy without a confusion matrix — overall accuracy hides per-class failure modes.

## Exercise 6 — tidymodels CV workflow

```{r}
set.seed(123)
split <- initial_split(peng %>% drop_na(), strata = species)
train <- training(split); test <- testing(split)

rec <- recipe(flipper_length_mm ~ body_mass_g + bill_length_mm + bill_depth_mm,
              data = train)
wf  <- workflow() %>% add_model(linear_reg()) %>% add_recipe(rec)

folds <- vfold_cv(train, v = 5, strata = flipper_length_mm)
res   <- fit_resamples(wf, folds, metrics = metric_set(rmse, rsq))
collect_metrics(res)
```

**Interpretation.** 5-fold CV gives a less optimistic RMSE than a single train/test split. Expect RMSE around 5-6 mm.

## Exercise 8 — Three-model bake-off

```{r, eval=FALSE}
spec_lm  <- linear_reg() %>% set_engine("lm")
spec_rf  <- rand_forest(trees = 500) %>% set_engine("ranger") %>% set_mode("regression")
spec_svm <- svm_rbf() %>% set_engine("kernlab") %>% set_mode("regression")

rec <- recipe(body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species,
              data = train) %>% step_dummy(all_nominal_predictors())

models <- workflow_set(
  preproc = list(simple = rec),
  models  = list(lm = spec_lm, rf = spec_rf, svm = spec_svm)
)

res <- workflow_map(models, resamples = folds, metrics = metric_set(rmse, rsq))
rank_results(res, rank_metric = "rmse")
```

**Interpretation.** On clean tabular data with strong species signal, linear regression often beats random forest — complexity is not free.

## Exercise 14 — Ridge vs OLS

```{r, eval=FALSE}
library(glmnet)
X <- model.matrix(body_mass_g ~ . - 1, data = peng %>% drop_na())
y <- peng %>% drop_na() %>% pull(body_mass_g)
cv <- cv.glmnet(X, y, alpha = 0)
coef(cv, s = "lambda.min")
```

**Interpretation.** Ridge shrinks all coefficients toward zero; lasso (alpha = 1) sets some exactly to zero. With only a handful of predictors, regularisation rarely beats OLS.

## Common student pitfalls

::: {.callout-tip}
- Reporting R² for a logistic regression (it doesn't mean what they think).
- Skipping the residual plots for linear models.
- Adding interactions without comparing to a main-effects model.
- Using accuracy as the sole classification metric when classes are imbalanced.
- Tuning hyperparameters on the test set instead of an inner CV loop.
- Calling tidymodels a "library" — it's a meta-package.
:::
