---
title: "Solutions — 09-advanced-modeling"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — DALEX variable importance on a random forest

```{r, eval=FALSE}
library(ranger)
library(DALEX)
library(palmerpenguins)

peng <- penguins %>% drop_na()
rf <- ranger(body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species + sex,
             data = peng, importance = "permutation")

ex <- explain(rf, data = peng %>% select(-body_mass_g), y = peng$body_mass_g,
              label = "ranger")
vi <- model_parts(ex)
plot(vi)
```

**Interpretation.** Permutation importance is the gold standard — it measures how much performance drops when each predictor is shuffled. Sex and species usually dominate; flipper length comes next.

## Exercise 2 — Partial dependence

```{r, eval=FALSE}
pdp_flipper <- model_profile(ex, variables = "flipper_length_mm")
plot(pdp_flipper)
```

**Interpretation.** PD curves show the marginal effect after averaging out other predictors. A monotonic rise here means heavier penguins have longer flippers on average — ecologically sensible. Flat PDs are diagnostic of a non-effect.

## Exercise 3 — SHAP for a single prediction

```{r, eval=FALSE}
shap <- predict_parts(ex, new_observation = peng[1, ], type = "shap")
plot(shap)
```

**Interpretation.** SHAP decomposes the prediction into per-feature contributions that sum to (prediction − baseline). Useful when a stakeholder asks "why did the model say X for this individual case?"

## Exercise 4 — Time-series decomposition

```{r}
data("AirPassengers")
dec <- decompose(AirPassengers, type = "multiplicative")
plot(dec)
```

**Interpretation.** Trend + seasonal + random. Use multiplicative when seasonal amplitude grows with the level, additive otherwise. The remainder should look like white noise — if it has structure, the decomposition missed something.

## Exercise 5 — ARIMA forecast

```{r, eval=FALSE}
library(forecast)
fit <- auto.arima(AirPassengers)
fc  <- forecast(fit, h = 24)
autoplot(fc)
```

**Interpretation.** Prediction intervals widen rapidly. Honest answer to "how far ahead": until the 95% PI becomes wider than the historical range. Beyond that the forecast is a guess.

## Exercise 6 — Phenology trend with autocorrelation

```{r, eval=FALSE}
library(nlme)
# Pretend dataset: arrival_doy by year
fit <- gls(arrival_doy ~ year, data = phen, correlation = corAR1())
summary(fit)
```

**Interpretation.** AR(1) errors absorb the year-to-year carryover that would otherwise inflate the trend's significance. Don't report a phenology trend without controlling for autocorrelation.

## Exercise 7 — Model agreement on importance

```{r, eval=FALSE}
# Compare ranger, xgboost, and a simple lm on the same data
# Compute importance for each, plot side by side.
```

**Interpretation.** If three different models disagree on the top variable, no model is trustworthy yet. Stable importance rankings across model families are a basic sanity check.

## Exercise 8 — Ensemble interpretation

```{r, eval=FALSE}
library(stacks)
# Build a stack of rf + glmnet + boost, blend predictions,
# then explain the blended model with DALEX as in Exercise 1.
```

**Interpretation.** Stacks usually pick small weights for weak models. The blended explanation often resembles whichever base model has the largest weight.

## Common student pitfalls

::: {.callout-tip}
- Treating variable importance as causal evidence.
- Forecasting beyond the horizon the prediction intervals support.
- Decomposing a series that isn't actually seasonal.
- Using SHAP for global explanations (it's per-prediction by design).
- Reporting only the highest accuracy across many tried models — selection-induced optimism.
:::
