---
title: "Solutions — 10-conservation"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(vegan)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.

## Exercise 1 — Shannon and Simpson diversity

```{r}
data(BCI)             # vegan: Barro Colorado Island tree counts
H <- diversity(BCI, index = "shannon")
D <- diversity(BCI, index = "simpson")
tibble(plot = seq_along(H), shannon = H, simpson = D) %>% head()
```

**Interpretation.** Shannon weights rare species more; Simpson weights common species more. The two indices usually rank communities similarly but disagree at the margins — that's when you report both.

## Exercise 2 — Linear vs non-linear population trend

```{r}
yrs <- 2000:2024
N   <- round(80 * exp(-0.04 * (yrs - 2000)) + rnorm(length(yrs), 0, 3))
df  <- tibble(year = yrs, abundance = N)

lin <- lm(abundance ~ year, data = df)
exp_fit <- nls(abundance ~ a * exp(b * (year - 2000)),
               data = df, start = list(a = 80, b = -0.04))

AIC(lin, exp_fit)
```

**Interpretation.** Compare AIC. Populations rarely decline linearly forever — exponential or logistic models reflect biological reality. Use the linear model only as a null benchmark.

## Exercise 3 — BACI design for a proposed protected area

A BACI (Before-After-Control-Impact) design needs:

- **Before / After:** pre-establishment surveys (≥ 3 years recommended) and post-establishment surveys (≥ 5 years to detect change above noise).
- **Control / Impact:** a paired unprotected site matched on habitat, elevation, and baseline community composition.
- **Response variables:** species richness, abundance of indicator species, vegetation structure.
- **Analysis:** mixed-effects model with `site_type * period` interaction. The interaction term is the BACI effect.

```{r, eval=FALSE}
library(lme4)
m <- lmer(richness ~ site_type * period + (1 | site), data = survey)
summary(m)
```

**Interpretation.** Without the control site, any post-establishment change is confounded with regional trends.

## Exercise 4 — Threat matrix

```{r}
threats <- tribble(
  ~threat,                ~severity, ~scope, ~irreversibility,
  "Agricultural expansion",       3,      3,                2,
  "Selective logging",            2,      3,                1,
  "Hunting pressure",             3,      2,                2,
  "Climate change",               3,      3,                3,
  "Invasive species",             2,      2,                2
) %>% mutate(score = severity * scope * irreversibility) %>%
  arrange(desc(score))

threats
```

**Interpretation.** Following IUCN's threat-classification logic, score = severity × scope × irreversibility. Climate change usually tops local matrices because it scores maximum on all three axes.

## Exercise 5 — One-page policy brief structure

A defensible brief contains:

1. **Headline finding** (one sentence, no jargon).
2. **What we did** (sampling design, sample size, time period).
3. **What we found** (effect size + uncertainty, not just p-values).
4. **What it means** (translated for the decision being made).
5. **What we recommend** (specific, actionable, ranked).
6. **What we don't know** (honest limitations).
7. **Contact + data availability**.

**Interpretation.** Mark down briefs that lead with methods or that hide uncertainty. The audience makes a decision regardless — your job is to make sure the decision is informed.

## Exercises 6-8 (sketch)

- **Complementarity for site selection** — use `prioritizr` to pick the smallest set of sites covering all species at least once.
- **Climate vulnerability** — combine exposure (projected climate change), sensitivity (thermal tolerance), and adaptive capacity (dispersal ability) into a composite index.
- **Community monitoring trends** — fit a GLM with year as predictor and report the annual rate of change with a confidence interval.

## Common student pitfalls

::: {.callout-tip}
- Reporting Shannon without checking that sample sizes are comparable.
- Calling a linear trend "the population is declining" when residuals show curvature.
- Designing BACI without a control site — that's just BA.
- Threat matrices with no explicit scoring rubric.
- Policy briefs longer than two pages or hiding effect sizes inside dense paragraphs.
:::
