---
title: "Solutions — 11-r-integrations"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

> **Instructor copy — do not distribute.** See `solutions/README.md`.
>
> Most exercises in this chapter are setup/configuration tasks. The
> answers below are what to look for when marking, not standalone
> code outputs.

## Exercise 1 — jamovi Syntax mode round-trip

What to look for:

- Student installed jamovi successfully (Flatpak on Linux, `.dmg`/Homebrew on macOS).
- Switched to Syntax mode and pasted the generated `jmv::ANOVA(...)` call.
- Reproduced identical F-statistic, df, and p-value in R.

```{r}
#| eval: false
library(jmv)
jmv::ANOVA(
  data = iris,
  dep = "Sepal.Length",
  factors = "Species",
  effectSize = "eta",
  postHoc = ~ Species
)
```

Common slip: students paste the call but forget `library(jmv)` and report a "could not find function" error.

## Exercise 2 — Positron vs RStudio

Acceptable observations include:

- VS Code-style keybindings instead of RStudio shortcuts.
- Variables pane shows both R and Python objects (no equivalent in RStudio).
- Console/Terminal split; Quarto preview pane embedded.
- Right-click "Run cell" works the same but addins that depend on `rstudioapi` are absent.

Reject pure aesthetic answers ("the icons are different").

## Exercise 3 — reticulate + scikit-learn

```{r}
#| eval: false
library(reticulate)
virtualenv_create("dains-py", packages = c("scikit-learn", "pandas"))
use_virtualenv("dains-py")

sk <- import("sklearn.linear_model")
fit <- sk$LinearRegression()$fit(r_to_py(mtcars[, c("wt", "hp")]), mtcars$mpg)
preds <- fit$predict(r_to_py(mtcars[, c("wt", "hp")]))

library(ggplot2)
ggplot(tibble(actual = mtcars$mpg, predicted = preds),
       aes(actual, predicted)) +
  geom_point() + geom_abline(slope = 1, linetype = 2) +
  theme_minimal()
```

Mark down: hard-coded system Python paths, missing virtualenv, or pulling Python objects into R without `py_to_r()` and then complaining about printing.

## Exercise 4 — plumber endpoint for penguin body size

```{r}
#| eval: false
# File: penguin_api.R
library(palmerpenguins)
library(dplyr)

#* @get /size
#* @param species The penguin species (Adelie, Chinstrap, Gentoo)
function(species) {
  penguins |>
    filter(species == !!species) |>
    summarise(mean_mass = mean(body_mass_g, na.rm = TRUE)) |>
    pull(mean_mass)
}
```

```bash
# Run:
Rscript -e 'plumber::pr("penguin_api.R") |> plumber::pr_run(port = 8000)'

# Test:
curl "http://localhost:8000/size?species=Adelie"
```

Common failure: forgetting `!!species` and ending up filtering by the literal string "species".

## Exercise 5 — IRkernel and Jupyter

What to look for:

- `IRkernel::installspec()` ran successfully and the R kernel appears in `jupyter kernelspec list`.
- Notebook reproduces a chapter 5 analysis (e.g., one-way ANOVA on `iris`) cell by cell.
- `quarto convert notebook.ipynb` produced a `.qmd` file with the same content.

Mark down: students who tried to install IRkernel from CRAN without also having Jupyter on the system PATH — that throws a confusing error.

## Exercise 6 — DuckDB + dbplyr

```{r}
#| eval: false
library(DBI)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb(), tempfile())
crops <- readr::read_csv("../data/agriculture/crop_yields.csv")
dbWriteTable(con, "crops", crops)

q <- tbl(con, "crops") |>
  filter(year >= 2010) |>
  group_by(country) |>
  summarise(mean_wheat = mean(wheat_yield, na.rm = TRUE)) |>
  arrange(desc(mean_wheat))

dbplyr::show_query(q)   # the generated SQL
collect(q)              # actually execute

dbDisconnect(con, shutdown = TRUE)
```

The SQL `show_query()` reveals should include `GROUP BY country` and `ORDER BY ... DESC`. Students who `collect()` first and then call `summarise()` are doing the analysis in R, not in the database, and have missed the point.

## Common student pitfalls

::: {.callout-tip}
- Installing tools via `pip` system-wide instead of `pipx` and breaking the system Python.
- Mixing conda, system Python, and reticulate virtualenvs in the same project.
- Assuming jamovi/JASP output is tidyverse-friendly — it isn't.
- Treating Positron's beta status as a blocker for coursework; for most R + Quarto workflows it's already fine.
- Hand-writing SQL in DBI when `dbplyr` would generate it from dplyr verbs.
:::
